---
title: "QC methods for DE analysis using DESeq2 - Answer key"
---

```{r data_setup}
#| echo: false

# load libraries needed to render this lesson
library(ggplot2)
library(DESeq2)

# load objects needed to render this lesson
dds  <- readRDS("./data/intermediate_dds.RDS")
meta <- readRDS("./data/metadata.RDS")
```

# Exercise 1

```{r pca_plot}
#| echo: false

# Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)

# Plot PCA 
plotPCA(rld, intgroup="sampletype")
```

**What does the above plot tell you about the similarity of samples?**

Samples from different experimental groups are different, while replicates within the same group are similar.

**Does it fit the expectation from the experimental design?**

Yes, it does.

**What do you think the %variance information (in the axes titles) tell you about the data in the context of the PCA?**

PC1 is associated with 37% of the variance in the data, and PC2 is associated with 16%. 

:::callout-note
You won't see the %variance information when you make PCA plots without the plotPCA() function. But there are tools that let you explore this.
:::