---
title: "Exploring DESeq2 results from the Wald test - Answer Key"
author:
  - Will Gammerdinger
date: "2025-11-13"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

# Exercise 1

**MOV10 Differential Expression Analysis: Control versus Knockdown**

Now that we have results for the overexpression results, do the same for the Control vs. Knockdown samples.

**Create a contrast vector called contrast_kd.**

```{r}
#| label: set_contrasts
#| eval: false

# Define contrasts for MOV10 knockdown
contrast_kd <- c("sampletype", "MOV10_knockdown", "control")
```

**Use contrast vector in the results() to extract a results table and store that to a variable called res_tableKD.**

```{r}
#| label: extract_results
#| eval: false

# Extract results for MOV10 knockdown vs control
res_tableKD <- results(dds, contrast=contrast_kd, alpha = 0.05)
```

**Shrink the LFC estimates using lfcShrink() and assign it back to res_tableKD.**

```{r}
#| label: apply_fold_change
#| eval: false

# Apply fold change shrinkage
res_tableKD <- lfcShrink(dds, coef="sampletype_MOV10_knockdown_vs_control", type="apeglm")
```