---
title: "Summarizing results from the Wald test - Answer key"
---

```{r data_setup}
#| echo: false

# load libraries needed to render this lesson
library(tidyverse)
library(DESeq2)

# load objects needed to render this lesson
res_tableOE <- readRDS("./data/intermediate_res_tableOE.RDS")
res_tableKD <- readRDS("./data/intermediate_res_tableKD.RDS")
padj.cutoff <- 0.05

# Create a tibble of results
res_tableOE_tb <- res_tableOE %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
# Subset the tibble to keep only significant genes
sigOE <- res_tableOE_tb %>%
  dplyr::filter(padj < padj.cutoff)
```

# Exercise 1

**MOV10 Differential Expression Analysis: Control versus Knockdown**

**Using the same p-adjusted threshold as above (`padj.cutoff < 0.05`), subset `res_tableKD` to report the number of genes that are up- and down-regulated in Mov10_knockdown compared to control.**

```{r}
# Create a tibble of results
res_tableKD_tb <- res_tableKD %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
```

**How many genes are differentially expressed in the Knockdown compared to Control? How does this compare to the overexpression significant gene list (in terms of numbers)?**

```{r}
# Subset the tibble to keep only significant genes
sigKD <- res_tableKD_tb %>%
  dplyr::filter(padj < padj.cutoff)
```

There are `rnrow(sigKD)` genes differentially expressed in the Knockdown compared to Control, and `r nrow(sigOE)` genes differentially expressed in the Overexpression compared to Control. Therefore, fewer genes are present in the Knockdown significant gene list.